---
title: "In-class Exercise"
editor: visual
---

# Overview

# Getting Started

Code chunk to install [sf](https://r-spatial.github.io/sf/) and tidyverse

```{r}
pacman::p_load(sf, tidyverse, spdep)

```

# Importing Geospatial Data

## Importing polygon feature dataframe

```{r}
mpsz<-st_read(dsn='data/geospatial',layer='MP14_SUBZONE_WEB_PL')
```

Geometry column in blob or list in R indicates if polygon, line or point.

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 15 fields in `mpsz` simple feature data frame. `mpsz` is in **svy21**projected coordinates systems. The bounding box provides the x extend and y extend of the data.

## Importing polyline feature data in shapefile form

```{r}
cyclingpath = st_read(dsn = "data/geospatial", layer = "CyclingPath")
```

The message above reveals that there are a total of 1625 features and 2 fields in `cyclingpath`linestring feature data frame and it is in **svy21** projected coordinates system too.

## Importing GIS data in kml format and transforming into projected coordinates system

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml") %>%
  st_transform(crs=3414)

```

The message above reveals that `preschool` is a point feature data frame. There are a total of 1359 features and 2 fields. Different from the previous two simple feature data frame, preschool is in **wgs84** coordinates system (long,lat).

Conversion to SVY21 using st_transform with code 3414.

# Reading Hunan dataframe

```{r}
hunan_sf = st_read(dsn = "data/geospatial", layer = "Hunan")
```

```{r}
hunan = read_csv("data/aspatial/Hunan_2012.csv")
```
